---
title: "covid19 - Portugal within the EU"
output: 
  flexdashboard::flex_dashboard:
    theme: cerulean
    orientation: rows
    navbar:
      - { title: "powered by Bruno A Lima",  href: "https://github.com/balima78/covid19data", align: right }

---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(ggbump)
library(plotly)
library(DT)

## dados
# ler dados ecdc: https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide
ecdc20201101 <- read_csv2("dados/2.preparada/COVID-19-geographic-disbtribution-worldwide-2020-11-01.csv") %>% 
  filter(GeoId %in% c("AT","BE","BG","CY","CZ","DE","DK","EE","EL","ES","FI","FR","HR","HU",
                      "IE","IT","LT","LU","LV","MT","NL","PL","PT","RO","SE","SI","SK","UK")) %>% 
  rename("country" = `Countries and territories`)

# informação demografica de: https://www.worldometers.info/population/countries-in-europe-by-population/
popEU <- read_csv2("dados/2.preparada/popEU.csv")

# calcular soma comulativa e filtrar resultados a partir de 2020-02-20
dados <- ecdc20201101 %>% arrange(country, DateRep) %>% 
  group_by(GeoId) %>%
  mutate(cum_cases = cumsum(Cases),
         cum_deaths = cumsum(Deaths),
         CFR = round(cum_deaths/cum_cases * 100,1)) %>% 
  filter(DateRep > "2020-02-20") 

# juntar data '21-02-2020' aos paises a que falta
dados<-bind_rows(dados,
             data.frame(
               DateRep = rep(as.Date("2020-02-21"),9),
               Day = rep(21,9),
               Month = rep(2,9),
               Year = rep(2020,9),
               Cases = rep(0,9),
               Deaths = rep(0,9),
               country = c("Portugal","Bulgaria","Cyprus","Hungary","Latvia","Malta","Poland","Slovakia","Slovenia"),
               GeoId = c("PT","BG","CY","HU","LV","MT","PL","SK","SI"),
               cum_cases = rep(0,9)
               )
)

# calcular casos comulativos por milhão, área e densidade
dados<-dados %>% left_join(popEU %>% select(GeoId,pop1m, landArea, density)) %>% 
  mutate(cum_cases_m = round(cum_cases/pop1m,3),
         cum_cases_a = round(cum_cases/landArea,3),
         cum_cases_d = round(cum_cases/density,3))

# coluna (dia) com numeração dos dias a partir de 10 casos
dados10<-dados %>% select(DateRep,GeoId, cum_cases) %>% 
  filter(cum_cases >=10) %>% arrange(GeoId,DateRep) %>% 
  group_by(GeoId) %>% mutate(dia = row_number()) %>% ungroup()
# join a dados
dados<-dados %>% left_join(dados10 %>% select(DateRep,GeoId, dia), by=c("DateRep","GeoId"))

# coluna (ordem) com rank diario de casos por milhao de habitantes
dadosOrd<-dados %>% arrange(DateRep,desc(cum_cases_m)) %>% 
  group_by(DateRep) %>% mutate(ordem = row_number()) %>% ungroup()
# coluna (ordem) com rank diario de casos por land Area
dadosOrd2<-dados %>% arrange(DateRep,desc(cum_cases_a)) %>% 
  group_by(DateRep) %>% mutate(ordem2 = row_number()) %>% ungroup()
# coluna (ordem) com rank diario de casos por densidade
dadosOrd3<-dados %>% arrange(DateRep,desc(cum_cases_d)) %>% 
  group_by(DateRep) %>% mutate(ordem3 = row_number()) %>% ungroup()

```

About
===================================

Row
-------------------------------------

### covid19 cases in Portugal
```{r}
valor<-max(
  dados %>% filter(GeoId=="PT") %>% .$cum_cases
)
valueBox(valor, icon = "fa-users", 
         color = "danger")

```

### Updated at
```{r}
hoje<-valor<-max(
  dados %>% filter(GeoId=="PT") %>% .$DateRep, na.rm = T
)
valueBox(hoje, icon = "fa-calendar-alt", 
         color = "warning")

```

Row {.tabset}
-------------------------------------
### Data - covid cases

Data for covid19 cases in the European Union (EU) + UK is updated from European Centre for Disease Prevention and Control (ecdc), available at:

https://www.ecdc.europa.eu/en/publications-data/download-todays-data-geographic-distribution-covid-19-cases-worldwide

This data is update each morning by the ecdc. Although, for Portugal, the most updated information is available at:

https://covid19estamoson.gov.pt/estatisticas/

A simple SIR model:

https://balima.shinyapps.io/sir_shiny/


Other Maps and visuals for covid19 are listed at:

http://www.cidrap.umn.edu/covid-19/maps-visuals


Open-Source-COVID-19:

http://open-source-covid-19.weileizeng.com/


### Demographic data

Demographic data in the table is available from: 

https://www.worldometers.info/population/countries-in-europe-by-population/

```{r}
popEU %>% select(-ln) %>% 
  datatable()
```

### Case Fatality Rate

$Case Fatality Rate (CFR, in \%)$ =  $\frac{Number of deaths from disease}{Number of diagnosed cases}$ x $100$

```{r}
dados.hoje=dados %>% filter(DateRep == hoje)
cfr<-ggplot(dados.hoje,
       aes(x = cum_cases, y = cum_deaths, label = country, size = CFR)) + 
  geom_point(color = ifelse(dados.hoje$GeoId == "PT", "red", "grey50")) +
  scale_y_log10() +
  scale_x_log10() +
  labs(x = "number of diagnosed cases",
         y = "number of deaths", 
         title = "Number of deaths by the number of positive cases (log10 transform)") +
    theme_bw()

ggplotly(cfr)
```

### 5 days incident cases
```{r}
dados.new<-dados %>% group_by(GeoId) %>% 
  mutate(Cases.1 = lag(Cases, order_by = DateRep),
         Cases.2 = lag(Cases, n=2, order_by = DateRep),
         Cases.3 = lag(Cases, n=3, order_by = DateRep),
         Cases.4 = lag(Cases, n=4, order_by = DateRep),
         Cases.5 = lag(Cases, n=5, order_by = DateRep)) %>% 
  ungroup() %>% mutate(newCases = Cases + Cases.1 + Cases.2 + Cases.3 + Cases.4 + Cases.5)

glag<-ggplot(dados.new, aes(cum_cases, newCases, color = country)) + 
  geom_line() + 
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  scale_y_log10() +
  scale_x_log10() +
  labs(x = "cumulative cases",
         y = "incident cases in 5 days window", 
         title = "Number of new cases in 5 days by cumulative cases (log10 transform)") +
  theme_bw()

ggplotly(glag)
```

### Incident cases
```{r}
g0<-ggplot(dados, aes(DateRep,Cases, color = country)) + 
  geom_point() + geom_line() +
  theme_bw() + 
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "Incident cases", color = "Country",
       title = "Evolution of incident cases")

  ggplotly(g0)
```



Cases per million inhabitants
==========================================

Row {.tabset}
-------------------------------------

### rank of cases

```{r}
# selecionar as colunas a utilizar
df<-dadosOrd %>% select(DateRep,ordem, GeoId)
#grafico
ggplot(df, aes(DateRep, ordem, color = GeoId)) +
  geom_point(size = 2) +
  geom_text(data = df %>% filter(DateRep == min(DateRep)),
            aes(x = DateRep - .5, label = GeoId), size = 3, hjust = 1) +
  geom_text(data = df %>% filter(DateRep == max(DateRep)),
            aes(x = DateRep + .5, label = GeoId), size = 3, hjust = 0) +
  geom_bump(size = 2, smooth = 8) + theme_bw() +
  # scale_x_continuous(limits = c(2010.6, 2013.4),
  #                    breaks = seq(2011, 2013, 1)) +
  #theme_minimal_grid(font_size = 14, line_size = 0) +
  theme(legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Rank for cases / million inhabitants",
       x = NULL,
       title = "Bump chart with the daily rank of cases by million inhabitants") +
  scale_y_reverse(breaks = seq(1,28, 1)) +
  scale_color_manual(breaks = c("AT","BE","BG","CY","CZ","DE","DK","EE","EL","ES","FI","FR","HR","HU",
                                "IE","IT","LT","LU","LV","MT","NL","PL","PT","RO","SE","SI","SK","UK"), 
                     values = c("#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6",
                                "#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","red","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6"))
  

```

### cumulative evolution
```{r}
# casos por milhao de habitantes
g1<-ggplot(dados, aes(DateRep,cum_cases_m,color = country)) + 
  geom_point() + geom_line() +
  theme_bw() + 
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "cum Cases / million inhabitants", color = "Country",
       title = "Evolution of daily cumulative cases /  million inhabitants")

  ggplotly(g1)
```

### Log10 transformation
```{r}
g2<-ggplot(dados, aes(DateRep,cum_cases_m, color = country)) + 
  geom_point() +
  geom_line() +  
  scale_y_continuous(trans='log10') +
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "log10 transform - cum Cases / million inhabitants", color = "Country",
       title = "Evolution of daily log10 transformation of cumulative cases /  million inhabitants") 

ggplotly(g2)
```

### Scatter plot
```{r}
source("fxs.R")

s1<-scat(data = dados %>% filter(DateRep == hoje),
     xx = pop1m,
     lab.x = "Population (millions)",
     tit = paste("Cases by country's population at",hoje)
)

ggplotly(s1)
```


Cases per Land Area
==========================================

Row {.tabset}
-------------------------------------

### rank of cases

```{r}
# selecionar as colunas a utilizar
df<-dadosOrd2 %>% select(DateRep,ordem2, GeoId)
#grafico
ggplot(df, aes(DateRep, ordem2, color = GeoId)) +
  geom_point(size = 2) +
  geom_text(data = df %>% filter(DateRep == min(DateRep)),
            aes(x = DateRep - .5, label = GeoId), size = 3, hjust = 1) +
  geom_text(data = df %>% filter(DateRep == max(DateRep)),
            aes(x = DateRep + .5, label = GeoId), size = 3, hjust = 0) +
  geom_bump(size = 2, smooth = 8) + theme_bw() +
  # scale_x_continuous(limits = c(2010.6, 2013.4),
  #                    breaks = seq(2011, 2013, 1)) +
  #theme_minimal_grid(font_size = 14, line_size = 0) +
  theme(legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Rank for cases / land Area",
       x = NULL,
       title = "Bump chart with the daily rank of cases by land Area") +
  scale_y_reverse(breaks = seq(1,28, 1)) +
  scale_color_manual(breaks = c("AT","BE","BG","CY","CZ","DE","DK","EE","EL","ES","FI","FR","HR","HU",
                                "IE","IT","LT","LU","LV","MT","NL","PL","PT","RO","SE","SI","SK","UK"), 
                     values = c("#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6",
                                "#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","red","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6"))
  

```

### cumulative evolution
```{r}
# casos por Area
g3<-ggplot(dados, aes(DateRep,cum_cases_a, color = country)) + geom_point() + geom_line() +
  theme_bw() +   
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "cum Cases / Land Area", color = "Country",
       title = "Evolution of daily cumulative cases /  land Area")

ggplotly(g3)
```

### Log10 transformation
```{r}
g4<-ggplot(dados, aes(DateRep,cum_cases_a, color = country)) + 
  geom_point() + 
  geom_line() +  
  scale_y_continuous(trans='log10') +
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "log10 transform - cum Cases / Land Area", color = "Country",
       title = "Evolution of daily log10 transformation of cumulative cases /  land Area")

ggplotly(g4)
```

### Scatter plot
```{r}
s2<-scat(data = dados %>% filter(DateRep == hoje),
     xx = landArea,
     lab.x = "Land Area (km^2)",
     tit = paste("Cases by country's land area at",hoje)
)

ggplotly(s2)
```

Cases per populational density
==========================================

Row {.tabset}
-------------------------------------

### rank of cases

```{r}
# selecionar as colunas a utilizar
df<-dadosOrd3 %>% select(DateRep,ordem3, GeoId)
#grafico
ggplot(df, aes(DateRep, ordem3, color = GeoId)) +
  geom_point(size = 2) +
  geom_text(data = df %>% filter(DateRep == min(DateRep)),
            aes(x = DateRep - .5, label = GeoId), size = 3, hjust = 1) +
  geom_text(data = df %>% filter(DateRep == max(DateRep)),
            aes(x = DateRep + .5, label = GeoId), size = 3, hjust = 0) +
  geom_bump(size = 2, smooth = 8) + theme_bw() +
  # scale_x_continuous(limits = c(2010.6, 2013.4),
  #                    breaks = seq(2011, 2013, 1)) +
  #theme_minimal_grid(font_size = 14, line_size = 0) +
  theme(legend.position = "none",
        panel.grid.major = element_blank()) +
  labs(y = "Rank for cases / density",
       x = NULL,
       title = "Bump chart with the daily rank of cases by density") +
  scale_y_reverse(breaks = seq(1,28, 1)) +
  scale_color_manual(breaks = c("AT","BE","BG","CY","CZ","DE","DK","EE","EL","ES","FI","FR","HR","HU",
                                "IE","IT","LT","LU","LV","MT","NL","PL","PT","RO","SE","SI","SK","UK"), 
                     values = c("#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6",
                                "#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","red","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6","#A9A6A6"))
  

```

### cumulative evolution
```{r}
# casos por densidade populacional
g5<-ggplot(dados, aes(DateRep,cum_cases_d, color = country)) + 
  geom_point() + 
  geom_line() +
  theme_bw() +
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "lcum Cases / density", color = "Country",
       title = "Evolution of daily cumulative cases /  density")

ggplotly(g5)
```

### Log10 transformation
```{r}
g6<-ggplot(dados, aes(DateRep,cum_cases_d, color = country)) + 
  geom_point() + 
  geom_line() +  
  scale_y_continuous(trans='log10') +
  scale_color_manual(values = ifelse(unique(dados$country) == "Portugal","red","#A9A6A6")) +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  labs(y = "log10 transform - cum Cases / density", color = "Country",
       title = "Evolution of log10 tranformation of daily cumulative cases /  density")

ggplotly(g6)
```

### Scatter plot
```{r}
s3<-scat(data = dados %>% filter(DateRep == hoje),
     xx = density,
     lab.x = "Country's density (population / km^2)",
     tit = paste("Cases by country's density at",hoje)
)

ggplotly(s3)
```
